babette: BEAUti, BEAST and Tracer for R

### 2018-02-06 article.tex
Gone through it all

  \textbf{1. }
    In the field of phylogenetics, 
     BEAST2 is one of the most widely used software tools. 
     It comes with the graphical programs BEAUti 2 and Tracer to 
     facilitate the creation of input files to BEAST2, and
     to interpret BEAST 2's created output files. 
     However, when many input files are needed, 
     a workflow of GUI programs is cumbersome. 
     Moreover, many other phygenetics tools are available 
     in R, which requires switching from one platform 
     to the other. \\
  \textbf{2. }
    Here, we present a free, libre and open-source package, \verb;babette;, 
    'BEAUti, BEAST and Tracer for R', for the R programming language. 
    \verb;babette; creates BEAST2 input files, runs BEAST2 and parses its results, 
    from an R function call. \\
  \textbf{3. }
    We describe \verb;babette;'s usage, the novel functionality it provides
    compared to BEAUti, and give some examples. \\
  \textbf{4. }
    As \verb;babette; is designed to be of high quality and extendable, 
    we conclude by describing the further development of the package \\
\end{abstract}

@@

Here, to provide such functionality we present \verb;babette;, 
’BEAUti, BEAST and Tracer for R’, which creates BEAST2 configuration files 
from an R function call and hence
will save time and tedious mouse clicking and 
reduces the chances of errors in such repetitive actions.
The interface of \verb;babette; mimics BEAUti. This
familiarity helps both beginner and experienced BEAST2 users 
to create configuration files using \verb;babette;.
Because there are many R packages for exploring the 
resulting phylogenetic trees, 
\verb;babette; enables the creation of a single-script 
pipeline from sequence alignments to tree analysis in R.

\verb;babette; has a similar goal as \verb;BEASTmasteR; \cite{beastmaster} (still
in pre-release), as both
allow for scripted use of BEAST2. The difference is that \verb;babette; 
has its focus on DNA alignments and ultrametric trees, 
where \verb;BEASTmasteR; is used for 
morphologucal traits and tip-dating.

@@

\verb;babette; is written in the R programming language \cite{R}
and creates a BEAST2 configuration file from an R function call,
in a similar way that BEAUti does.

\verb;babette;'s main function is \verb;create_beast2_input_file;, which, 
as the name suggests, creates a BEAST2 configuration file. 
\verb;create_beast2_input_file; needs at least the name of a 
FASTA file containing a DNA alignment
and a name for the to-be-created configuration file. 
The default settings for the other arguments of \verb;create_beast2_input_file; 
are identical to BEAUti's default settings.
Per alignment, a site model, clock model and tree prior can be chosen.
Multiple alignments can be used, each with its own (unlinked) site model, clock model and tree prior.

\verb;babette; currently has 61 exported functions to create a BEAST2 configuration file. 
\verb;babette; is an alternative for a majority of BEAUti use cases, 
but does not yet support the full functionality of BEAUti. 
Because of BEAUti's high number of plugins, 
\verb;babette; uses a software architecture that expects te be extended.

\verb;beastier; \cite{beastier} and \verb;lumier; \cite{lumier} are related packages, used to confirm the correct working of verb;babette;.
\verb;lumier; calls BEAST2 from within R and is used to confirm that the XML files created by \verb;babette; are valid. 
Additionally, \verb;lumier; is used to run BEAST2 to create posteriors. 
\verb;beastier; can parse BEAST2 posteriors, and is used to confirm whether posteriors have an estimated or fixed crown age.

@@

\begin{lstlisting}[language=R, caption=Loading, label=lst:loading_babette, floatplacement=H]
library(babette)
library(beautier)
library(beastier)
library(tracerer)
\end{lstlisting}

BEAUti, and likewise \verb;babette;, needs at least a FASTA filename
and an output filename. In BEAUti, this is achieved by loading a FASTA file, 
then saving an output file using a common
save file dialog. In \verb;babette;, the same is achieved by:

\begin{lstlisting}[language=R, caption=Simplest example, label=lst:simplest_example, floatplacement=H]
create_beast2_input_file(
  "alignment.fas",
  "beast2.xml"
)
\end{lstlisting}

This code will create a BEAST2 configuration file with name '\verb;beast2.xml;',
using a FASTA file with name \verb;alignment.fas;, using the same default settings as BEAUti.
The default settings are, among others, to use a Jukes-Cantor site model \cite{cantor1969mammalian}, 
a strict clock, and a Yule birth tree prior \cite{yule}.

@@

\begin{lstlisting}[language=R, caption=Example with different site model and clock model and tree prior, label=lst:all_different, floatplacement=H]
create_beast2_input_file(
  "alignment.fas",
  "beast2.xml",
  site_models = create_hky_site_model(),
  clock_models = create_rln_clock_model(),
  tree_priors = create_bd_tree_prior()
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, caption=Two alignments, label=lst:two_alignments, floatplacement=H]
create_beast2_input_file(
  c("anthus_aco.fas", "anthus_nd2.fas"),
  "beast2.xml",
  site_models = list(
    create_tn93_site_model(), 
    create_gtr_site_model()
  )
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, caption=Example with Yule tree prior with different birth rate distribution, label=lst:diff_distr, floatplacement=H]
library(babette)
create_beast2_input_file(
  "alignment.fas",
  "beast2.xml",
  tree_priors = create_yule_tree_prior(
    birth_rate_distr = create_exp_distr()    
  )
)
\end{lstlisting}

% TODO: Mix sentences better
Our initial motivation to create \verb;babette; 
is that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. It does not allow for fixing
the crown age. Without \verb;babette;, one needs to manually edit the BEAST2 
XML configuration file, which is tedious and prone to errors. 
Fixing the crown ages allows for theoretical experiments 
with one less source of variation.

@@

\begin{lstlisting}[language=R, caption=Example with fixed crown age, label=lst:fixed_crown_age, floatplacement=H]
create_beast2_input_file(
  "alignment.fas",
  "beast2.xml"
  posterior_crown_age = 15
)
\end{lstlisting}

Instead of creating a BEAST2 input file and then running BEAST2,
\verb;babette; allows to do both steps at once:

\begin{lstlisting}[language=R, caption=Run BEAST2, label=lst:run_beast2, floatplacement=H]
out <- run_beast2(
  "alignment.fas"
)
\end{lstlisting}

\verb;run_beast2; also allows for setting a site model, clock model,
tree prior and a fixed crown age. The variable \verb;out; 
is a list that contains the posterior
parameter estimates, posterior phylogenies (one per alignment)
and operator acceptances.

@@

\begin{lstlisting}[language=R, caption=Show posterior phylogenies, label=lst:densitree, floatplacement=H]
library(phangorn)
densiTree(out$anthus_aco_trees)
\end{lstlisting}

@@

\verb;babette; uses the Travis CI \cite{travis} 
continuous integration service, which is known to significantly 
increase the the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
\verb;babette; has a 100\% code coverage, which correlates with code quality \cite{horgan1994,del1995correlation}. 
\verb;babette; follows Hadley Wickham's style guide \cite{style_guide}, which improves software quality \cite{fang2001}.
\verb;babette; is dependent on multiple packages, which are 
\verb;APE; \cite{APE}, 
\verb;beastier; \cite{beastier},
\verb;devtools; \cite{devtools},
\verb;geiger; \cite{GEIGER},
\verb;ggplot2; \cite{ggplot2},
\verb;knitr; \cite{knitr},
\verb;lumier; \cite{lumier},
\verb;phangorn; \cite{phangorn},
\verb;rmarkdown; \cite{rmarkdown},
\verb;seqinr; \cite{seqinr},
\verb;stringr; \cite{stringr},
\verb;testit; \cite{testit} and 
\verb;TreeSim; \cite{TreeSim}.

@@

A roadmap of \verb;babette;'s future extensions can be found on its GitHub.

>>>>

  \textbf{1. }
    In the field of phylogenetics, 
     BEAST2 is one of the most widely used software tools. 
     It comes with the graphical programs BEAUti 2 to
     facilitate the creation of input files to BEAST2, and
     with DensiTree and Tracer 
     to interpret BEAST 2's created output files. 
     However, when many input and output files are required, 
     a workflow of GUI programs is cumbersome. 
     Moreover, many other phygenetics tools are available 
     in R, which requires switching from one platform 
     to the other. \\
  \textbf{2. }
    Here, we present a free, libre and open-source package, \verb;babette;, 
    'BEAUti 2, BEAST2 and Tracer for R', for the R programming language. 
    \verb;babette; creates BEAST2 input files, runs BEAST2 and parses its results, 
    from an R function call. \\
  \textbf{3. }
    We describe \verb;babette;'s usage, the novel functionality it provides
    compared to BEAUti 2, and give some examples. \\
  \textbf{4. }
    As \verb;babette; is designed to be of high quality and extendable, 
    we conclude by describing the further development of the package \\
\end{abstract}

@@

BEAST2 is also associated with Tracer \cite{tracer} and 
DensiTree 2 \cite{DensiTree}. Both are desktop applications 
to analyze the output of BEAST2, each with a user-friendly graphical user interface.

Tracer's purpose is to analyze the parameter estimates generated
from a BEAST 2 run. It shows, among
others, the effective sample size (ESS) and time series ('the trace', 
hence the name) of each variable in the MCMC run. Both ESS and trace
are needed to assess the strength of the inference. 
DensiTree visualizes the phylogenies of a BEAST2 posterior, with
many options to improve the displayal of those many phylogenies.

However, for exploring the output of many BEAST2 runs, 
one would like a script to collect all parameter ESSes,
parameter traces and posterior phylogenies.

Here, to provide such functionality we present \verb;babette;, 
’BEAUti 2, BEAST2 and Tracer for R’, 
which creates BEAST2 configuration files,
runs BEAST2 and analyzes its results 
from an R function call and hence
will save time and tedious mouse clicking and 
reduces the chances of errors in such repetitive actions.
The interface of \verb;babette; to create BEAST2 configureation
files closely mimics BEAUti. This
familiarity helps both beginner and experienced BEAST2 users 
to create configuration files using \verb;babette;.
\verb;babette; enables the creation of a single-script 
pipeline from sequence alignments to tree analysis in R.

\verb;babette; is the first R package that unifies the
full workflow of working with BEAST2. An
example to create a BEAST2 input file is 
\verb;BEASTmasteR; \cite{beastmaster}.
Also \verb;BEASTmasteR; allows 
to create BEAST2 configuration files from R. 
The difference is that \verb;babette; 
has its focus on DNA alignments and ultrametric trees, 
where \verb;BEASTmasteR; is used for 
morphological traits and tip-dating.
Examples of R packages to parse the BEAST2 output files
are \verb;rBEAST; \cite{rBEAST}, 
\verb;RBEAST; \cite{RBEAST} and 
\verb;rbeast2; \cite{rbeast2}.

@@

\verb;babette; is written in the R programming language \cite{R}
and enables the full BEAST2 workflow from an R function call,
in a similar way that BEAUti, DensiTree and Tracer do.

\verb;babette;'s main function is \verb;run_beast2;, which, 
configures BEAST2, runs it and parses its output. 
\verb;run_beast2; needs at least the name of a 
FASTA file containing a DNA alignment. 
The default settings for the other arguments of \verb;run_beast2; 
are identical to BEAUti's and BEAST2's default settings.
Per alignment, a site model, clock model and tree prior can be chosen.
Multiple alignments can be used, each with its own (unlinked) site model, 
clock model and tree prior.

\verb;babette; currently has 61 exported functions to set up  
a BEAST2 configuration file. 
\verb;babette; is an alternative for a majority of BEAUti use cases, 
but does not yet support the full functionality of BEAUti. 
Because of BEAUti's high number of plugins, 
\verb;babette; uses a software architecture that expects te be extended.

\verb;babette; has 7 exported function to run and help run BEAST2.
One function is a wrapper function to run BEAST2, others
allow the user to check if a BEAST2 configuration file is indeed valid.

\verb;babette; has 20 exported function to parse the BEAST2 output
file and analyse the created posterior. \verb;babette; gives the
same ESSes and summary statistics as Tracer. The data is formatted
as such, that it can easily be visualized using ggplot2 (for a trace,
similar to Tracer)
or phangorn (for the phylogenies in a posterior, simular to DensiTree).

@@

\begin{lstlisting}[language=R, caption=Loading, label=lst:loading_babette, floatplacement=H]
library(babette)
\end{lstlisting}

BEAUti, and likewise \verb;babette;, needs at least a FASTA filename
to produce a BEAST2 configuration file. 
In BEAUti, this is achieved by loading a FASTA file, 
then saving an output file using a common
save file dialog. After this, BEAST2 needs to be invoked
on the created configuration file to create multiple files
storing the created posterior. Finally, these output
files must be parsed by either Tracer of DensiTree.

In \verb;babette;, the same is achieved by:

\begin{lstlisting}[language=R, caption=Simplest example, label=lst:simplest_example, floatplacement=H]
out <- run_beast2(
  "alignment.fas"
)
\end{lstlisting}

This code will create a (temporary) BEAST2 configuration file,
using a FASTA file with name \verb;alignment.fas;, using the same default settings as BEAUti.
\verb;babette; will then execute a BEAST2 with that file, and
parses the output. The default settings are, among others, 
to use a Jukes-Cantor site model \cite{cantor1969mammalian}, 
a strict clock, and a Yule birth tree prior \cite{yule}.

@@

\begin{lstlisting}[language=R, caption=Example with different site model and clock model and tree prior, label=lst:all_different, floatplacement=H]
out <- run_beast2(
  "alignment.fas",
  site_models = create_hky_site_model(),
  clock_models = create_rln_clock_model(),
  tree_priors = create_bd_tree_prior()
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, caption=Two alignments, label=lst:two_alignments, floatplacement=H]
out <- run_beast2(
  c("anthus_aco.fas", "anthus_nd2.fas"),
  site_models = list(
    create_tn93_site_model(), 
    create_gtr_site_model()
  )
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, caption=Example with Yule tree prior with different birth rate distribution, label=lst:diff_distr, floatplacement=H]
out <- run_beast2(
  "alignment.fas",
  tree_priors = create_yule_tree_prior(
    birth_rate_distr = create_exp_distr()    
  )
)
\end{lstlisting}

Our initial motivation to create \verb;babette; 
is that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. It does not allow for fixing
the crown age. Without \verb;babette;, one needs to manually edit the BEAST2 
XML configuration file, which is tedious and prone to errors. 
Fixing the crown ages allows for theoretical experiments 
with one less source of variation.

@@

\begin{lstlisting}[language=R, caption=Example with fixed crown age, label=lst:fixed_crown_age, floatplacement=H]
out <- run_beast2(
  "alignment.fas",
  posterior_crown_age = 15
)
\end{lstlisting}

@@

\begin{lstlisting}[language=R, caption=Show posterior phylogenies, label=lst:densitree, floatplacement=H]
densitree(out$anthus_aco_trees)
\end{lstlisting}

@@

\verb;babette; uses the Travis CI \cite{travis} 
continuous integration service, which is known to significantly 
increase the the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
\verb;babette; has a 100\% code coverage, which correlates with code quality \cite{horgan1994,del1995correlation}. 
\verb;babette; follows Hadley Wickham's style guide \cite{style_guide}, which improves software quality \cite{fang2001}.
\verb;babette; is dependent on multiple packages, which are 
\verb;APE; \cite{APE}, 
\verb;beautier; \cite{beautier},
\verb;beastier; \cite{beastier},
\verb;devtools; \cite{devtools},
\verb;geiger; \cite{GEIGER},
\verb;ggplot2; \cite{ggplot2},
\verb;knitr; \cite{knitr},
\verb;phangorn; \cite{phangorn},
\verb;rmarkdown; \cite{rmarkdown},
\verb;seqinr; \cite{seqinr},
\verb;stringr; \cite{stringr},
\verb;testit; \cite{testit} and 
\verb;tracerer; \cite{tracerer},
\verb;TreeSim; \cite{TreeSim}.

@@
